---
title: "data_import"
output: github_document
---
This document will show how to import data 



```{r setup,echo=FALSE, message = FALSE }
library(tidyverse)
library(readxl)
library(haven)
```

# Import the FAS Litters CSV 

Import the dataset and clean the names
```{r}
litters_df= read_csv("data_import_examples/FAS_litters.csv")
view(litters_df)

names(litters_df)

litters_df = janitor::clean_names(litters_df)
names(litters_df)
```

## Import the FAS Pups CSV 

Import the dataset and clean the names

Use relative paths for collaborations

```{r}
pups_df= read_csv("data_import_examples/FAS_pups.csv")
view(pups_df)

names(pups_df)

pups_df = janitor::clean_names(pups_df)
names(pups_df)
```


# Look at read_csv options 

col_names and Skip very first rows 

```{r}
litters_df= 
  read_csv(
    file= "data_import_examples/FAS_litters.csv",
      col_names= FALSE, 
      skip= 2
  )
```

Handling Missing Data. Helps R understand missing columns with missing values are not character values if there are numbers in it. Helps distinguish between character and numbers variables. 

```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv",
  na= c("NA","",".")
  )
litters_df= janitor::clean_names(litters_df)

pull(litters_df, gd0_weight)
```

What if we code 'group' as a factor variable?

```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv",
  na= c("NA","","."),
  col_types = cols(
    Group = col_factor()
    )
  )
litters_df
```

# Import an excel file 

Import MLB 2011 summary data 
```{r}

mlb_df= read_excel("data_import_examples/mlb11.xlsx", sheet ="mlb11")
mlb_df
  
```

## Import SAS data

```{r}
pulse_df= read_sas("data_import_examples/public_pulse_data.sas7bdat")
pulse_df
```

## Never use read.csv() 

```{r}
litters_df = read.csv("data_import_examples/FAS_litters.csv")
litters_df
```

Never do this either: $ 
```{r}
litters_df$L 
```



# Look at the dataset

# Section 1