---
title: "Data_Manipulation"
author: "Kimberly Lopez"
date: "2024-09-19"
output: github_document
---
This document will show *data manipulation*



```{r setup,echo=FALSE, message = FALSE }
library(tidyverse)
library(readxl)
library(haven)
```

Import the two datasets that we're going to manipulate. Used Janitor function to clean the names in each df. 

```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv",
  na= c("NA","",".")
  )

litters_df= janitor::clean_names(litters_df)

pups_df= 
  read_csv("data_import_examples/FAS_pups.csv",
  na= c("NA","",".")
  ) 
pups_df= janitor::clean_names(pups_df)

```

## 'select' 

Use 'select()' to select variables. Selects which column you want to work with.

```{r}
select(litters_df, group, litter_number, gd0_weight)

```

Select a range of columns between group column and gd18_weight. 

```{r}
select(litters_df, group:gd18_weight)
```

To get rid of columns you dont want to use: 
```{r}
select(litters_df, -pups_survive) 
```

To get rid of 
```{r}
select(litters_df, -(group:gd18_weight))
```

To select columns that start with the same letters like "gd" you can select by:
```{r}
select(litters_df, starts_with("gd"))
```

To selct columns that contain key phrases 
```{r}
select(litters_df, contains("pups"))
```
Use select to rename variables
```{r}
select(litters_df, GROUP=group)

rename(litters_df, GROUP= group) 

select(litters_df, litter_number, gd0_weight, everything())

```

Use relocate function to reorder the variables 
```{r}
relocate(litters_df, litter_number, gd0_weight)
```
Example: In the pups data, select the columns containing litter number, sex, and PD ears
```{r}
select(pups_df, litter_number, sex, pd_ears)
```


## 'filter'

This applies a condition which variables meet it will keep. This helps answer questions about a variable. 

- Double equalls asks "is this variable =" 
```{r}
filter(litters_df,gd_of_birth ==20)
filter(litters_df, gd_of_birth == 19)
```

- If you want variables > than a # use: 
```{r}
filter(litters_df, pups_born_alive >8)
filter(litters_df, pups_born_alive >=8)
```

- Not equal to: 
```{r}
filter(litters_df, pups_born_alive != 9 )
```

# For Character Vectors: 
 - for if it is exactly equal to one condition 
  - for if it meet either or two conditions using %in% "is in"

```{r}
filter(litters_df, group == "Low8")
filter(litters_df, group %in% c("Low 7", "Low8"))
```

- for if it meets either or two conditions in *two variables/ columns* 
```{r}
filter(litters_df, group %in% c("Low 7", "Low8"), pups_born_alive ==8 )
```

- To get rid of rows with *missing values* 
- dont use is.na instead use drop_na() this drops rows with any missing data 

```{r}

drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```
Example: Pups 
In the pups data:

- Filter to include only pups with sex 1
- Filter to include only pups with PD walk less than 11 and sex 2
```{r}
filter(pups_df, sex== 1)
filter(pups_df, pd_walk < 11, sex ==2)
```


## 'mutate'

To subtract an entire row by another row and defined a new variable: 
- subtract weight gain and gd0 weight gain to get new variable wt_gain
- can square values to get rid of skewness

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)

```
 
 - can operate on character variables 
 - overwriting on a string variable to be all lower case inside of a df 
 - manipulate multiple things at the same time too
 
```{r}
mutate(litters_df, group = str_to_lower(group))

mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight, 
  group = str_to_lower(group)
)
```
 
 In the pups data:

- Create a variable that subtracts 7 from PD pivot
- Create a variable that is the sum of all the PD variables
```{r}
mutate(pups_df, pivot_minus7 = pd_pivot - 7)

mutate(pups_df, pd_sum = pd_ears + pd_eyes + pd_pivot + pd_walk)
```

## 'arrange' 

Normally for aestheics, 
- can reorder by weight default 
- can reorder by descending value 
- arrange by multiple things as the same time 
```{r}

arrange(litters_df, gd0_weight)

arrange(litters_df, desc(gd0_weight))

arrange(litters_df, pups_born_alive, gd0_weight)
```
## Piping 

Use piping when doing multiple data maniuplations at the same time. Use |>

DO THIS 
```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() |> 
  mutate(
    wt_gain = 
      gd18_weight - gd0_weight,
    group = str_to_lower(group)
  ) |> 
  select(-pups_born_alive) |> 
  filter(group == "Con7") 
```

Data export. Write csv
```{r}
write_csv(litters_df, "data_import_examples/cleaned_fas_litters.csv")
```


