---
title: "tidy_data"
author: "Kimberly Lopez"
date: "2024-09-24"
output: html_document
---

Load packages 
```{r}
library(tidyverse)
library(haven)
```


```{r}
pulse_df= 
  read_sas("data_import_examples/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()

```

Go from wide format to long format. State columns and where they will be stores given respective values. Use this if multiple variables contain the same values for a type of data. 
```{r}
pulse_tidy_df= 
  pulse_df |>
  pivot_longer(
    cols= bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to= "bdi")
pulse_tidy_df

```

In the bdi column, there are prefix of bdi_score we can strip off the values by: 
```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_df
```

After, we should mutate the visit variable to replace values "bl" since it is not in the format of months. Replace values "b" to 00m: 
- relocate the columns so visit is in the second columns 
```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")|>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) |>
  relocate(id,visit)

print(pulse_tidy_df, n = 12)
```

To do all the coding in one shot: 
- janitor
- pivot wide to long 
- remove common prefix 
- mutate data & relocate 
```{r}
pulse_df = 
  haven::read_sas("data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) |>
  relocate(id, visit)

print(pulse_df, n = 12)
```

Doing one more example with the Litters data: 
- gd_weight has 2 columns in the original dataset
- use case match function to automatically take a variable and recode/replace as a value. Must list out all the options. 

```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv")|>
  janitor::clean_names() |>
  pivot_longer(
    gd0_weight:gd18_weight, 
    names_to = "gd_time",
    values_to = "weight"
  )|>
  mutate(
    gd_time = case_match(
      gd_time, 
      "gd0_weight" ~0,
      "gd18_weight" ~ 18, 
      
    )
  )
litters_df
```


```{r}

  
```


