---
title: "tidy_data"
author: "Kimberly Lopez"
date: "2024-09-24"
output: github_document
---

Load packages 
```{r}
library(tidyverse)
library(haven)
```


```{r}
pulse_df= 
  read_sas("data_import_examples/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()

```

# Pivot longer
Go from wide format to long format. State columns and where they will be stores given respective values. Use this if multiple variables contain the same values for a type of data. 
```{r}
pulse_tidy_df= 
  pulse_df |>
  pivot_longer(
    cols= bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to= "bdi")
pulse_tidy_df

```

In the bdi column, there are prefix of bdi_score we can strip off the values by: 
```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_df
```

After, we should mutate the visit variable to replace values "bl" since it is not in the format of months. Replace values "b" to 00m: 

- relocate the columns so visit is in the second columns 
```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")|>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) |>
  relocate(id,visit)

print(pulse_tidy_df, n = 12)
```

To do all the coding in one shot: 

- janitor
- pivot wide to long 
- remove common prefix 
- mutate data & relocate 
```{r}
pulse_df = 
  haven::read_sas("data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) |>
  relocate(id, visit)

print(pulse_df, n = 12)
```

Doing one more example with the Litters data: 
- gd_weight has 2 columns in the original dataset
- use case match function to automatically take a variable and recode/replace as a value. Must list out all the options. 

```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv")|>
  janitor::clean_names() |>
  pivot_longer(
    gd0_weight:gd18_weight, 
    names_to = "gd_time",
    values_to = "weight"
  )|>
  mutate(
    gd_time = case_match(
      gd_time, 
      "gd0_weight" ~0,
      "gd18_weight" ~ 18, 
      
    )
  )
litters_df
```


Lets make up an analysis result table.
```{r}
analysis_df= 
  tibble(
    group = c("treatment","control","control","treatment"),
    time = c("pre","post","pre","post"),
    mean = c(4,10,4.2,5)
    
  )
  
```

# Pivot Wider
Pivot Wider for readability. Helpful for post analysis results. 

- took previous dataset 
- took column names from time variable 
- values are coming from the mean variable
- use knitr::kable() for r built in readability

```{r}
analysis_df |>
  pivot_wider(
    names_from = time, 
    values_from = mean
  )|>
  knitr::kable()
```

# Binding Rows 

Load data, to create a tidy dataframe that has all the variables from 3 different tables.

Binding Tables

- read excels with ranges that correspond to each table in the data set 
- ensure each table corresponds with their movie name using mutate()

```{r}
fellowship_ring = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")

```

Use Bind rows to stack the tables together 

- also clean up using janitor 
- pivot to longer 

```{r}
lotr_df= 
  bind_rows(fellowship_ring, two_towers,return_king)|> 
  janitor:: clean_names()|>
  pivot_longer(
    female:male,
    names_to = "sex", 
    values_to = "words"
    ) |>
  relocate(movie)|>
  mutate(race = str_to_lower(race))


```


# Joining Datasets 

There are four major ways join dataframes x and y:
- Inner: keeps data that appear in both x and y
- Left: keeps data that appear in x
- Right: keeps data that appear in y
- Full: keeps data that appear in either x or y

Join FAS datasets 

- Import "litters" dataset
- mutates so that we know what weight gained 
- seperate the variable 
- use seperate to seperate column 
```{r}
litters_df= 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA","","."))|>
  janitor::clean_names()|>
  mutate( 
    wt_gain = gd18_weight - gd0_weight,
   )|>
  separate(
    group, into = c("dose","day_of_treatment"), sep=3 
  )
litters_df
```
Import FAS "pups" dataset 

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv",na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    sex = as.factor(sex))

pups_df
```
 Join the datasets 
- matching by variable litter number
- using left joining into the pups data frame. 
- rearrange so litter_number, dose, day of treatment follow order

```{r}
fas_df = 
  left_join(pups_df, litters_df, by = "litter_number")|>
  relocate (litter_number, dose, day_of_treatment)

fas_df
```

# Other Functions on Names 
- left_join(x, y)
- inner_join (, )
* if there are multiple of the same matched data, it will give warnings 
- anti_join(, )


